---
title: "More Practice with the ggformula Package"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---
  
```{r setup, include=FALSE}
library(learnr)
library(tibble)
library(ggformula)
library(ggthemes)
library(mosaic)
library(palmerpenguins)
library(readr)
library(NHANES)
data("NHANES", package = "NHANES")
theme_set(theme_bw())
D <- data.frame(
  group = LETTERS[1:3],
  count = c(20, 25, 18)
)
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  fig.width = 6, fig.height = 2.5)
```


<style type="text/css">
  span.boxed {
    border:5px solid gray;
    border-radius:10px;
    padding: 5px;
  }
span.invboxed {
  border:5px solid gray;
  padding: 5px;
  border-radius:10px;
  color: white;
}

table, td, th { border:0px; }

/* cellpadding */
  th, td { padding: 5px; }

</style>
  

## Data Sets

This tutorial is designed to give you some more practice with `ggformula` 
and to introduce a few more features of the system. Mostly you will get 
a chance to explore on your own and try things out.

Here are four data sets you can use in this tutorial.

* `HELPrct` has data from a study of people addicted to alcohol, cocaine, or heroine
* `KidsFeet` has information about some kids' feet.
* `penguins` has some information about some penguins.
* `NHANES` has lots of variables for 10,000 people in a health study.

To find out more about the data sets use `?HELPrct`, `?KidsFeet`, `?penguins`,
or `?NHANES`.
To see the first few rows of the data, you can use `head()`.

```{r, head, exercise = TRUE}
head(penguins)
```

Another useful command for looking at the data is `glimpse()`. It will show the
first few values for each variable (like head does) but it flips this around so
each variable is on a row, which often works better on a screen or piece of
paper.

To get a list of functions available in `ggformula`, run this code chunk.

```{r, apropos, exercise = TRUE}
# list all functions starting gf_
apropos("gf_")
```

#### Your Turn

Make some plots to explore one or more of these data sets.

* Experiment with different types of plots.
* Use mapping and/or facets to reveal groups.
* You can put more than one plot in a code chunk, but we've provided two chunks
in case you want to separate your work that way.  Use one chunk for experimenting
and copy and paste your favorites to the other chunk if you like.

```{r, explore, exercise = TRUE}
gf_point(length ~ width, data = KidsFeet)
head(KidsFeet)
?KidsFeet
```

```{r, explore-more, exercise = TRUE}

```

Post your favorite plot(s) here by making a new slide in this google presentation:

 * [Favorite Plots Google Slides](https://docs.google.com/presentation/d/16wu5IVPkXtuCAybGbF5vxF3GQ_h4xqU3syFL3S6Gkq4/edit?usp=sharing)
 
You may work in a group of 2--3, but then you must submit as many plots
as there are people in your group.

## More Customization

You may be wondering how to have more control over 
things like:

* the colors, shapes, and sizes chosen when mapping attributes
* labeling, including titles and axis labels
* fonts, colors, and sizes of text
* color of plot elements like background, gridlines, facet labels, etc.

As you can imagine, all of these things can be adjusted pretty much however you
like.  We will cover a few of the most common options here.

## Custom axis labels

One of the most common customizations you will want to make to your plots will
be to change the title, subtitle, and axis labels (and maybe add a caption).
All these things can be done by chaining (`%>%`) the function `gf_labs()` with a
plot layer.

Check out the example below, and try changing the text labels to ones that make
sense to you. Note that all the input arguments to `gf_labs` are optional. So,
for example, you could alter *only* the x-axis label by chaining the command
`gf_labs(x='My X Axis Label')` with your plot.

```{r gf-labs-example, exercise=TRUE}
gf_bar(~Race1, data = NHANES) %>%
  gf_labs(title = "Race in NHANES Data",
    subtitle = "(2009-2012)",
    caption = "These data were collected by the US National Center for Health Statistics (NCHS), which has conducted a series of health and nutrition surveys since the early 1960's.",
    x = "",  # empty quotes here results in no x-axis label!
    y = "Number of Observations"
  )
```

## Custom axis limits

One way to zoom in (or out) is to filter the dataset so that only the data you
wish to plot is included.

You can also use `gf_lims()` to set custom x and y axis limits. 

1. Try different axis limits to see how the plot changes.
2. What happens if you set the min or max value for an axis to NA? (Hint: compare the plot to a plot with *no* custom axis limits.)
3. What happens if you set the axis limits to be c(max, min) instead of c(min,max)?

```{r axis-limits, exercise=TRUE}
gf_point(births ~ day_of_year, data = Births1978) %>%
  gf_lims(x = c(100, 200), y = c(8000, 9000))
```

## Axis Labels that Don't Fit

Sometimes - particularly for bar graphs of categorical variables with long category names - axis tick labels overlap in an ugly way. For example:

```{r, echo=TRUE, fig.width=4}
gf_bar( ~ Education, data = NHANES)
```

### Flip the Axes

If you have already made your plot and notice that there isn't room
for the labels, one simple solution to solve this problem is to flip 
the x and y axes of the plot.

```{r, coord-flip, exercise = TRUE, echo=TRUE}
gf_bar( ~ Education, data = NHANES) %>%
  gf_refine(coord_flip())
```

If you design you plot that way from the beginning,
you can just flip the formula around.  But there is a catch: 
**Formulas must always have a right hand side**, so we need to
put a dot on the right side as an empty placeholder:

```{r, native-horizontal, exercise = TRUE, echo=TRUE}
gf_bar(Education ~ ., data = NHANES) 
```

### Rotate the Labels

Another solution is to rotate the axis labels. We can do that by modifying the `angle` and `hjust` values for the x tick labels in the plot's `theme`.  `angle` is the angle in degrees by which to rotate the labels, and `hjust` moves them up and down (positive `hjust` moves down, and negative moves up). For example:

```{r, rotate-labels, exercise = TRUE, echo=TRUE}
gf_bar( ~ Education, data = NHANES) %>% 
    gf_theme(axis.text.x = element_text(angle = 65, hjust = 1))
```

Try adjusting `angle` and `hjust` to see how those arguments work.

### Your Turn!
The dataset at <http://sldr.netlify.com/data/MammalMetabolicRates.csv> provides data on mammal metabolic rates.  Read it in and make a bar graph of the number of observations per Order (or per Family, Genus, or Species) with legible axis tick labels.

```{r mmr, exercise=TRUE, fig.height = 5}
Metabolic <- 
  read_csv('http://sldr.netlify.com/data/MammalMetabolicRates.csv')
Metabolic
```

