---
title: "CIs and PIs for Simple Linear Regression"
output: 
  learnr::tutorial:
    progressive: TRUE
    allow_skip: TRUE
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tibble)
library(ggformula)
library(mosaic)
# library(maxLik)
# library(car)
theme_set(theme_bw(base_size=16))
# knitr::opts_chunk$set(exercise.checker = checkr::checkr_tutor)
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  fig.width = 6, fig.height = 2.5)
tutorial_options(exercise.eval = FALSE)

set.seed(14)
bonobos <- read.csv('https://sldr.netlify.app/data/bonobo_faces.csv') %>%
  na.omit() %>%
  sample(size = 10)

earn <- foreign::read.dta('www/heights.dta') %>%
  mutate(sex = ifelse(sex == 2, 'female', 'male')) %>%
  na.omit() %>%
  dplyr::filter(earn > 0)

covid <- readr::read_csv('https://docs.google.com/spreadsheets/d/e/2PACX-1vTEjMAq1V8JVVFmEHtIg2NGBp5VMktqn8vauEviwxjR4em7Lcg0MG8Ts4p8wqwrK6Be44lMjm2AG9S6/pub?gid=0&single=true&output=csv')

covid <- covid %>%
  mutate(day = c(1:nrow(covid)),
         log2cases = log2(cases),
         case_diff = c(NA, diff(covid$cases)),
         norm_case_diff = diff(c(0, covid$cases)) / 
           c(1, head(covid$cases, -1)),
         case_ratio = covid$cases / c(NA,head(covid$cases, -1)))  %>%
  dplyr::filter(day >= 40) %>%
  mutate(day = day - day[1])

dinos <- read.csv('https://sldr.netlify.app/data/Campione_Evans_2012_data.csv')
```

## Learning Outcomes

In this module, we will build (just a little) on our tools to do *inference* on the simple linear regression model, devising methods to make *predictions* from a model *with uncertainty* (in other words, *predictions with confidence intervals*).

By the end of the module you will:

1) Be able to compute confidence intervals and prediction intervals for fitted values from a linear regression
2) Be competent to explain the difference between confidence and prediction intervals

*These materials come from your text book, FAStR2 Chapter 6.4.*

## Confidence and Prediction Intervals

In the last module, we learned to generate confidence intervals (or do hypothesis tests) for the slope and intercept parameters of a simple linear regression.

But often, we are interested in using fitted regression models to *predict* response variable values in specific scenarios. In other words, we're sometimes more interested in the predicted response variable values $\hat{y}$ than the parameter estimates $\beta_0$ and $\beta_1$.

It's a simple matter to compute point estimates of these fitted values: if we have an x-value of interest (let's call it $x_*$), we just compute $\hat{\beta}_0 + \hat{\beta}_1x_*$ to get our estimate.

But what about uncertainty? 

### Ways to Conceptualize Uncertainty in Predictions
*Text book reference: FAStR Chapter 6.4.7*

We could do this two ways!

- **A confidence interval:** We could find an interval that gives plausible values for "where the line is located", taking into account uncertainty in the intercept and slope estimates. Another way of thinking about this is that it gives a plausible range of values for the *expected value* (mean) of the response variable for all hypothetical new observations with $X = x_*$.
- **A prediction interval:** We could find and interval that gives plausible values for "the point cloud" of possible new observations, taking into account uncertainty in the slope and intercept, plus the residual variance. Another way of thinking about this is that it gives a plausible range of values for the response for one new datapoint with $X = x_*$.  Note that we've coined the new name *prediction interval* to differentiate it from the other one, this is really also a CI (just for a very specific scenario).

### Practical Example and Calculations

In order to compute these confidence and prediction intervals, we'll definitely need appropriate standard error estimates for the predicted values $\hat{y}$.  But before deriving them, let's first review a practical example just to keep the big picture front and center.

*Note: this video is from a great contemporary statistician and teacher, [Mine Çetinkaya-Rundel](http://www2.stat.duke.edu/~mc301/) - she's awesome...*

![](https://www.youtube.com/watch?v=qVCQi0KPR0s){ width=400px }

(You can also [watch directly on YouTube](https://www.youtube.com/watch?v=qVCQi0KPR0s) if you prefer.

## SE for a Confidence Interval
*Text book reference: FAStR Chapter 6.4.7*

If our goal is a **confidence interval** about $\hat{y}$, then we need to examine the sampling distribution of 

$$ \hat{Y}_* = \hat{\beta}_0 + \hat{\beta}_1x_*$$
or (again using $B_1$ and $B_2$ to reinforce the thought that the parameter estimates can be thought of as random  variables)

$$ \hat{Y}_* = B_0 + B_1x_*$$

Since $(\bar{x}, \bar{Y})$ is on the line, $\bar{Y} = B_0 + B_1\bar{x}$ and we can substitute  $\bar{Y} - B_1\bar{x}$ for $B_0$:

$$ \hat{Y}_* = \bar{Y} - B_1\bar{x} + B_1x_*$$

$$= \bar{Y} + B_1(x_* - \bar{x})$$

We already know that 

$$\bar{Y} \sim \text{Norm}(\beta_0 + \beta_1\bar{x}, \frac{\sigma}{\mid \mathbf{1}\mid})$$

and from our expression for the SE of $B_1$ (since $x_*$ and $\bar{x}$ are constants here),

$$ B_1(x_* - \bar{x}) \sim \text{Norm}(\beta_1x_* - \beta_1\bar{x}, \frac{\sigma(x_* - \bar{x})}{\mid \mathbf{x} - \mathbf{\bar{x}}\mid})$$

And we know that $\bar{Y}$ and $B_1$ are independent, so the variance of their sum is just the sum of their variances, and:

$$\hat{Y_*} \sim \text{Norm}\bigg{(}\beta_0 + \beta_1x_*, \sigma \sqrt{\frac{1}{\mid \mathbf{1} \mid ^2} + \frac{(x_* - \bar{x})^2}{\vert \mathbf{x} - \mathbf{\bar{x}}\vert^2}}\bigg{)}$$

Practically (substituting our sample statistics as parameter estimates and skipping the vector notation), we will estimate the standard error of this sampling distribution by $s \sqrt{\frac{1}{n} + \frac{(x_* - \bar{x})^2}{(n-1)s_x^2}}$ (like you saw in the video).

## SE for a Prediction Interval

The result we just got will help us to find the SE for a prediction interval, since it already takes into account two of the three sources of uncertainty we need to account for (the slope and intercept and residuals).

We'll start by considering the random variable $Y_* \sim \text{Norm}(\beta_0 + \beta_1x_*, \sigma)$ which models a "new" (that is, indepdendent of the data used to fit the model) observation of the response for $x = x_*$.

What is the sampling distribution of $Y_* - \hat{Y}_*$? Its standard error will help us compute the margin of error for our prediction interval. Luckily, again we can use the fact that the variance of a sum (or difference) is the sum of the variances, finding that

$$Y_* - \hat{Y}_* \sim \text{Norm}\bigg{(}0, \sigma \sqrt{1 + \frac{1}{\mid \mathbf{1} \mid ^2} + \frac{(x_* - \bar{x})^2}{\vert \mathbf{x} - \mathbf{\bar{x}}\vert^2}}\bigg{)}$$

So we can estimate the SE for a prediction interval via  $ s \sqrt{1 + \frac{1}{n} + \frac{(x_* - \bar{x})^2}{(n-1)s_x^2}}$ (again, like you saw in the video).

## R Shortcuts
Of course, it's tedious to compute these estimates by hand (and you already saw a brief example of how to do it in R in the video).

To recap, you can compute confidence or prediction intervals in R using `predict()`. Assuming you have a fitted model `m` with response `y` and predictor `x_star` in dataset `d`, and you want an interval estimate of the fitted value for $x = x_*$:

```{r, echo = TRUE, eval = FALSE}
m <- lm(y ~ x, data = d)
new_data <- data.frame(x = x_star)
conf_int <- predict(m, newdata = new_data, 
        interval = 'confidence',
        level = 0.95)

pred_int <- predict(m, newdata = new_data, 
        interval = 'prediction',
        level = 0.95)
```

If you wanted to make interval estimates for more than one value $x_*$, you can just enter a list of values in `new_data`:

```{r, echo = TRUE, eval = FALSE}
new_data <- data_frame(x = c(x_star1, x_star2, ...))
```

Finally, if you want to plot one or both intervals on a scatter plot of the data, you can!

So you can actually see the result, let's demonstrate with a regression predicting bonobo dominance score `normDS` by `weight` from the `bonobos` dataset.

```{r, echo = TRUE}
gf_point(normDS ~ weight, data =bonobos) %>%
  gf_lm(interval = 'prediction', color = 'black', 
        fill = 'seagreen') %>%
    gf_lm(interval = 'confidence') 
```

```{r pred-width}
quiz(
  question("In the plot above, which of the intervals is the prediction interval?",
           answer("The narrower one"),
           answer("The wider one", correct = TRUE, message = "Yes! The prediction interval will always be wider because it takes into account one additional source of uncertainty; or, in other words, the plausible range of values for one new observation will always be wider than that for the mean of a set of new observations."),
           answer("Impossible to tell without plotting them individually"),
           random_answer_order = TRUE,
           allow_retry = TRUE),
  question("If you wanted to use this model to compute a 95% CI for the dominance score of Kuni, a bonobo at the Jacksonville Zoo who weighs 36.7 kg, what kind of interval should you use?",
           answer("Confidence", message = 'Not quite; we want a prediction for a single new observation, not for the expected value of a *class* of new observations'),
           answer("Prediction", correct = TRUE,
                  message = "Yep! This is an interval estimate for a single new observation, so a prediction interval is appropriate."),
           allow_retry = TRUE),
  question("If you wanted to use this model to compute an interval estimate for the dominance score of 41kg male bonobos, what kind of interval should you use?",
           answer("Confidence", message = 'Yes! we want a prediction for the expected value of a *class* of new observations, so a confidence interval is the right choice.', correct = TRUE),
           answer("Prediction", message = "Not quite. This is an interval estimate for the mean of a class of new observations, so a prediction interval is not appropriate."),
           allow_retry = TRUE)
)
```

## Practice

Find an interval estimate for the assertiveness score (`AssR`) of Zeke, and 8-year-old male bonobo, based on the dataset `bonobos` (which also contains the variable `Age`).

```{r zeke-pred, exercise = TRUE}

```

```{r zeke-pred-hint-1}
m <- lm()
new_data <- data.frame()
predict()
```

```{r zeke-pred-hint-2}
m <- lm(AssR ~ Age, data = bonobos)
new_data <- data.frame()
predict()
```

```{r zeke-pred-hint-3}
m <- lm(AssR ~ Age, data = bonobos)
new_data <- data.frame(Age = 8)
predict()
```

```{r zeke-pred-hint-4}
m <- lm(AssR ~ Age, data = bonobos)
new_data <- data.frame(Age = 8)
predict(m, newdata = new_data, interval = ...)
```

```{r zeke-pred-hint-5}
m <- lm(AssR ~ Age, data = bonobos)
new_data <- data.frame(Age = 8)
predict(m, newdata = new_data, interval = 'prediction')
```

## More Practice

The dataset `earn` contains data on people's income (`earn`) and their height in inches (`height`) from a 1994 survey of Americans.  Fit a model predicting income as a function of height and then produce an interval estimate for the income of people who are 6'4" (76 inches) tall.

```{r tall-income, exercise = TRUE}

```

```{r tall-income-hint-1}
m <- lm()
new_data <- data.frame()
predict()
```

```{r tall-income-hint-2}
m <- lm(earn ~ height, data = earn)
new_data <- data.frame()
predict()
```

```{r tall-income-hint-3}
m <- lm(earn ~ height, data = earn)
new_data <- data.frame(height = 76)
predict()
```

```{r tall-income-hint-4}
m <- lm(earn ~ height, data = earn)
new_data <- data.frame(height = 76)
predict(m, newdata = new_data, interval = ...)
```

```{r tall-income-hint-5}
m <- lm(earn ~ height, data = earn)
new_data <- data.frame(height = 76)
predict(m, newdata = new_data, interval = 'confidence')
```

