---
title: "Plots with the ggformula package (NHANES)"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---
  
```{r setup, include=FALSE}
library(learnr)
# library(checkr)
# library(statPREP)
library(tibble)
library(ggformula)
library(ggthemes)
library(mosaic)
library(NHANES)
library(readr)
theme_set(theme_bw())
data(NHANES, package = "NHANES")
D <- data.frame(
  group = LETTERS[1:3],
  count = c(20, 25, 18)
)
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  fig.width = 6, fig.height = 2.5)
```


<style type="text/css">
  span.boxed {
    border:5px solid gray;
    border-radius:10px;
    padding: 5px;
  }
span.invboxed {
  border:5px solid gray;
  padding: 5px;
  border-radius:10px;
  color: white;
}

table, td, th { border:0px; }

/* cellpadding */
  th, td { padding: 5px; }

</style>
  
## Acknowledgements

This tutorial was created based closely on the "Introduction" tutorial provided with the `ggformula` package. 

This tutorial assumes you are already familiar with the basics of 
`ggformula`, perhaps from the tutorial about births in the United States.


## NHANES data

Time for some new data, some review, a few new wrinkles (like bar charts).

The `NHANES` dataset contains measurements from 10,000 human subjects in
the National Health and Nutrition Evaluation Survey.  To learn more about the data,
try one or more of these:

  * `?NHANES`
  * `names(NHANES)`
  * `glimpse(NHANES)`
  * `inspect(NHANES)`

```{r, NHANES, exercise = TRUE}
?NHANES
```

Because `NHANES` has so many more (and more kinds of) variables, 
we can create even more kinds of plots with `NHANES`.

## Histograms

### The basics

A histogram can help you visualize the distribution of one **quantitative** variable.

Histograms are bit different from scatter plots:

  * Instead of using `gf_point()`, we will use `gf_histogram()`.  (You probably
  guessed that.)
  
  * We only need to tell R about one variable (the x-variable).  R will calculate the 
  information for the y-variable for us!
  
  * So we use a formula with only one variable in it:  `~ x`.  (Notice that `x` goes on 
  the right side.)
  
### Practice with histograms

1. Try the example below as-is by hitting "Run Code".

2. Now experiment with plotting different **quantitative** variables from the `NHANES`
dataset. (Remember you can use `?` or `names(NHANES)` or
`glimpse(NHANES)` to get the variable names and more information.)

3. You can change the size of the bins using either `bins` (the number of bins)
or `binwidth` (the width of the bins).  Experiment with different bins.  (If you
don't provide `bins` or `binwidth` information, R will just make something up.
You can usually do better if you take control.)

4. To get density instead of counts on the y-axis, switch from function
`gf_histogram()` to `gf_dhistogram()`.


```{r hist, exercise=TRUE}
gf_histogram( ~ BPSysAve, data = NHANES, bins = 30)
```

## Density plots

A density plot is a smoothed contour tracing the shape of a dataset's
distribution.  The `gf_density()` and `gf_dens()` functions produce these plots
(in slightly different ways).  To compare with a histogram apples-to-apples, we
will use a **density histogram** (`gf_dhistogram()`).

<!-- [^dhistogram] -->

<!-- [^dhistogram]: A density histogram is rescaled so that the **total area** -->
<!-- of the histogram is **always 1**. -->

```{r, hist-dens, exercise = TRUE}
gf_dhistogram( ~ BPSysAve, data = NHANES) 
gf_dens( ~ BPSysAve, data = NHANES) 
```

#### Your turn

 1. Use `%>%` to overlay the two layers.

 2. Use `size = 2` or `size = 3` to make the density curve thicker.
 
 3. Change `gf_dens()` to `gf_density()`.
 
 4. Can you make the density (use `gf_density()`) plot a different color and
 partly transparent?  That will make it easier to see the histogram.
 
 5. Replace `BPSysAve` with a different quantitative variable.

## Boxplots and violin plots

Boxplots and violin plots provde a quick comparison of the distribution of 
a quantitative variable in different groups.

### Give it a try

1. Run the code below to see how this works (You're plotting systolic blood
pressure for seven groups who watch different amounts of TV each day).

2. Change `gf_boxplot()` to `gf_violin()` and see what happens.

3. Try changing the order of `x` and `y` in the formula -- what happens?

4. If you are interested, try different variables. Remember, `y` must be
quantitative (that is what you'll make boxplots of) and `x` must be categorical
(that defines the groups).

5. Change `gf_boxplot()` to `gf_point()`.  What happens?  Is this a useful plot?

6. Now try `gf_jitter()`.  Can you figure out what this does?  (You might like to 
set `alpha` to a smallish number.)


```{r, boxplot-groups, exercise=TRUE}
gf_boxplot(BPSysAve ~ TVHrsDay, data = NHANES)
```

Note: If you haven't seen these plots before, we will talk more about what they are
later.

### Just one variable?

Both `gf_boxplot()` and `gf_violin()` require a **two-sided formula**, so if you want to plot just one quantitative variable you have to replace the $x$ in the formula with empty quotes "".

```{r, boxplot, exercise=TRUE}
gf_boxplot( BPSysAve ~ "", data = NHANES)
```

Boxplots are usually most useful to compare the distribution of a variable
between several groups.  (If we have only one variable, we can use histograms
or density plots instead.)

## Bar graphs

Bar graphs help visualize the distribution of a **categorical** variable, 
and we can create them with `gf_bar()`. 

```{r, bar-graphs, exercise=TRUE}
gf_bar( ~ TVHrsDay, data = NHANES)
```

### Percents and Proportions?
What if we want to show the percent or proportion in each category, rather than the number of observations? `gf_percents()` and `gf_props()` to the rescue! Try changing the function from `gf_bar()` to `gf_percents()` or `gf_props()` and see what happens.

You can also add an 'h' at the end of each function to make the plots 'horizontal' with `gf_barh()`, `gf_propsh()`, and `gf_percentsh()`. Try that too!


```{r, prop-graphs, exercise=TRUE}
gf_bar( ~ TVHrsDay, data = NHANES)
```

```{r, prop-graphs-hint-1}
gf_props( ~ TVHrsDay, data = NHANES)
```

```{r, prop-graphs-hint-2}
gf_percents( ~ TVHrsDay, data = NHANES)
```

```{r, prop-graphs-hint-3}
gf_percentsh( ~ TVHrsDay, data = NHANES)
```



### Your Turn

 1. Create a bar graph for a different (categorical) variable.
 2. What happens if you use a quantitative variable?

```{r, more-bar-graphs, exercise=TRUE}
gf_bar( ~ TVHrsDay, data = NHANES)
```

## Bar graphs by groups

What if you want to show a set of proportions for each of several groups?  For
example, the NHANES data also includes a variable `Marijuana` that indicates
whether the person has used marijuana.  Is there a relationship?

```{r, echo=TRUE}
gf_bar( ~ TVHrsDay | Marijuana, data = NHANES)
```

## Stacked bar graphs

What if, instead of one figure panel per group, you want to see a stacked bar
graph for the same data?  Here's an example.  You use the input 

  * `fill= ~ variable_name` to specify the name of the variable that defines the groups
(here, `Marijuana`).

```{r, echo=TRUE}
gf_bar( ~ TVHrsDay, fill = ~ Marijuana, data = NHANES)
```

## Side-by-Side bar graphs

What if, instead of stacked bars, you want side-by-side bars? 
Simply add the additional argument

  * `position='dodge'`.

```{r, echo=TRUE}
gf_bar( ~ TVHrsDay, fill = ~ Marijuana, data = NHANES, position = 'dodge')
```

### Other groups?

Try to make a similar plot to one or more of the bar graph variations shown
above, but using a variable other than `Marijuana` to specify the groups. First,
we will look at the `head()` of the dataset to see what variables are there -
then it's up to you to choose one to plot, and modify the code above to make the
figure.

```{r, not-mari, exercise=TRUE}
head(NHANES)
```

## Bar graph with pre-tabulated data

Sometimes, you may be given data that is already tabulated. Instead of a dataset with one row for every case, you will have one row for every group, and a variable that gives the number of observations in each group.  For example, the simple dataset below is pre-tabulated:

```{r, echo=FALSE}
D <- data.frame(
  group = LETTERS[1:3],
  count = c(20, 25, 18)
)
D
```

We can use the function `gf_col` to make a bar graph of pre-tabulated data.
This function always expects the counts as the `y` part of the formula, and the
group names as `x` (after the tilde).

```{r col-plot, exercise=TRUE}
gf_col(count ~ group, data = D)
```


## Multiple layers with %>%

We said that `gf_point()` creates a plot with points.
This isn't quite true. Technically, it creates a **layer** with points.
A single plot may have multiple layers. For example, you might want a scatter plot with a trend line overlaid on it, or a histogram with a standard normal curve overlaid.

To create a multi-layered plot, simply
append `%>%` at the end of the code for one layer and follow that with
another layer. *(The `%>%` symbol is called a "pipe" because it sends the results of one operation on to the next operation for further processing)*

#### Exercise

1. If you run the following code as is, you will get two separate plots.
2. Combine these two layers into a single plot by appending `%>%` at
the end of the first line.
3. Try adding another layer for a third variable.


```{r multi-layers, exercise = TRUE}
gf_density(~ BPSysAve, data = NHANES)
gf_density(~ BPDiaAve, data = NHANES, fill = "blue")
```

<!-- #### Linear regression line -->

<!-- Adding a (simple) linear regression line to a scatter plot is a relatively common task, and ggformula provides an easy way to do it with the function `gf_lm()`. -->

<!-- For example, we might investigate whether there is a linear relationship between total cholesterol and weight using the NHANES data: -->

<!-- ```{r, lm-example, exercise=TRUE} -->
<!-- gf_point(TotChol ~ Weight, data = NHANES, alpha = 0.1) %>% gf_lm() -->
<!-- ``` -->


## Facets (Multi-panel plots)

If we want to look at all 20 years of birth data, overlaying the data is likely
to put too much information in too little space and make it hard to tell which
data is from which year.  (Even with good color and symbol choices, deciphering
20 colors or 20 shapes is hard.) Instead, we can put each year in separate
**facet** or sub-plot.  By default the coordinate systems will be shared across
the facets which can make comparisons across facets easier.

There are two ways to create facets.  The simplest way is to add a vertical bar `|` to
our formula.


```{r, births-facets, exercise = TRUE, fig.height = 6}
gf_point(births ~ day_of_year | year, data = Births, size = 0.5)
```

The second way is to add on a facet command using `%>%`:

```{r, births-facets-wrap, exercise = TRUE, fig.height = 6}
gf_point(births ~ day_of_year, data = Births, size = 0.5) %>%
gf_facet_wrap( ~ year)
```

#### Practice with facets

Edit the plot below to:

1. make one facet for each day of the week (`wday`)
2. map color to `year`

```{r, births-facets-ex1, exercise = TRUE, fig.height = 6}
gf_point(births ~ day_of_year, data = Births, 
         size = 0.5, color = "blue")
```

```{r births-facets-ex1-hint}
gf_point(births ~ day_of_year | wday, data = Births, 
         size = 0.5, color = ~ year)
```

## Facet Grids and Facet Wraps 

The faceting we did on the previous page is called facet wrapping.  If the facets
don't fit nicely in one row, the facets continue onto additional rows.

A facet grid uses rows, or columns, or both in a fixed way.

Notice that after the `|` there is now a *formula* instead of a single variable.  Try generating the plot below - can you figure out what the formula does?  If you need a hint, try changing `year ~ wday` to `wday ~ year` and see what happens... 

```{r, facet-grid, exercise = TRUE, fig.height = 9}
gf_point(births ~ day_of_year | year ~ wday, data = Births, size = 0.5)
```

## The facet grid formula 

Hopefully, you figured out that the facet grid formula is interpreted as "row
variable ~ column variable" -- the resulting plot will have one row of facets
for every value of the first variable, and one column of facets for every value
of the second variable.

#### Pracice with the facet grid formula

Recreate the plot below using `gf_facet_grid()`.  This works much like
`gf_facet_wrap()` and accepts a formula with one of three shapes

* `y ~ x` (facets along both axes)
* ` ~ x`  (facets only along x-axis)
* `y ~ .` (facets only along y-axis; notice the important dot in this one)  

(These three formula shapes can also be used on the right side of `|`.)


```{r, facet-grid2, exercise = TRUE, fig.height = 9}
gf_bar( ~ TVHrsDay | Marijuana ~ Gender, fill = ~ Marijuana, data = NHANES)
```

Feel free to try some other variables or other types of plots.

