---
title: "R Basics"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
# library(checkr)
# library(statPREP)
library(tidyverse)
library(ggformula)
library(mosaic)
theme_set(theme_bw())
# knitr::opts_chunk$set(exercise.checker = checkr::checkr_tutor)
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.width = 6, fig.height = 2.5)
MI_lead <- read.csv(file='http://sldr.netlify.com/data/MI_lead.csv')
tutorial_options(exercise.eval = FALSE)
```


<style type="text/css">
  span.boxed {
    border:5px solid gray;
    border-radius:10px;
    padding: 5px;
  }
span.invboxed {
  border:5px solid gray;
  padding: 5px;
  border-radius:10px;
  color: white;
}

table, td, th { border:0px; }

/* cellpadding */
  th, td { padding: 5px; }

</style>

## Your Mission
The purpose of this tutorial is to help you start to get familiar with the way R works, and some basic R commands.

But you may have noticed that *you are not even using R.* At least, you haven't logged in to the RStudio server.

This tutorial environment lets you read some helpful information, then practice writing and running your own R code, *without* tabbing back and forth between the tutorial website and RStudio. Here's hoping it provides a nice, gentle introduction in a controlled environment!

Tutorials can also serve as helpful reference guides later on, if you need reminders or more practice. We will keep using them to learn and practice R throughout the semester, and links to all our tutorials will be on the course Moodle page ("R Tutorials" section).

## Communicating with R
  You will do most of your work in R with *code* or *commands*. Instead of pointing and clicking, you will type one or more lines of code, which R will *execute* (doing the work you have asked it to do). Then, R will return the results of whatever operation you asked it to do - sometimes producing a plot, other times creating a plot. Sometimes executing code has almost no visible effect (no plot or text output is produced), but instead some object is created and stored in R's *environment* for later use.

  
### Two Key Questions

To get R (or any software) to do something for you,
there are two important questions you must be able to answer.  Before
continuing, think about what those questions might be.
  
### The Questions

To get R (or any software) to do a job for you, there are two
important questions you must be able to answer:
  
#### 1. What do you want the computer to do?
  
#### 2. What must the computer know in order to do that?
  
### Providing R with the information it needs
  
  R *functions* provide R with the answer to the first question: what do you want the computer to do?
  
  Most functions in R have short, but descriptive names that describe what they do.  For example, R has some functions to do basic mathematical operations:  the function `sqrt()` computes the square root of a number, and the function `round()` rounds a number (by default, it rounds to the nearest integer).
  
  But just giving R a function is not enough: you also need to answer the second question (what information does R need to do the job?). For example, if you want to use the function `round()`, you also need to provide R with the number you want to round!
  
  We will provide answers to our two questions by filling in
the boxes of a basic template:
  
  <center>
  <h2><strong><span class="boxed">function</span> ( <span class="boxed">&nbsp;information1&nbsp; , &nbsp;information2&nbsp; , ...)</strong> 
  </h2></center>
  &nbsp;

(The `...` indicates that there may be some additional *arguments* (input information we could provide to R) we could add
  eventually. Some functions need only one input, but if a function takes more than one argument, they are separated by commas.)

###Using simple functions
Let's practice what you just learned, trying out the `sqrt()` and `round()` functions. 

Edit the code below to compute the square root of 64:

```{r find-sqrt, exercise=TRUE}
function(information_R_needs)
```

```{r find-sqrt-hint-1}
sqrt(information_R_needs)
```

```{r find-sqrt-hint-2}
sqrt(64)
```

Now try computing the square root of 44, then rounding it to the nearest integer:

```{r round-sqrt, exercise=TRUE}
function1(information_R_needs)
function2(information_R_needs)
```

```{r round-sqrt-hint-1}
sqrt(information_R_needs)
function2(information_R_needs)
```

```{r round-sqrt-hint-2}
sqrt(44)
function2(information_R_needs)
```

```{r round-sqrt-hint-3}
sqrt(44)
round(information_R_needs)
```

```{r round-sqrt-hint-4}
sqrt(44)
round(sqrt(44)) # yes, you can nest functions. sometimes (not always) a good idea.
```

### Storing information in R: variables
In the last section, you computed the square root of 44 and then rounded it, probably like this:

```{r, echo=TRUE}
sqrt(44)
round(6.63325)
```

But to do that, you probably had to first find the root, make a note of the result, and then provide that number to the `round` function.  What a pain!

A better alternative, if you are computing a value that you will want to use later on, is to store it as a named variable in R.  In the previous example, you might want to store the square root of 44 in a variable called my_root; then you can provide my_root to the `round` function without checking the result of the `sqrt` calculation first:

```{r naming-variable, echo=TRUE}
my_root <- sqrt(44)
round(my_root)
```

Notice that to assign a name to the results of some R code, you use the symbol "<-". You can think of it as an *assignment arrow* -- it points from a result toward a name and assigns the name to the result.

Try editing the code to change the name of the variable from my_root to something else, then run your new code:

```{r name-a-variable, exercise=TRUE}
my_root <- sqrt(44)
round(my_root)
```

### What if I have a list of numbers to store?

Sometime you might want to create a variable that contains more than one number.  You can use the function `c` to *concatenate* a list of numbers:

```{r}
my_fave_numbers <- c(4, 44, 16)
my_fave_numbers
```

(First we stored the list of numbers, calling it my_fave_numbers; then we printed the results to the screen by simply typing the variable name my_fave_numbers).

Try making a list of your three favorite numbers, then using the function `sum` to add them all up:

```{r c-and-sum, exercise=TRUE}
```

```{r c-and-sum-hint, results='hide'}
my_numbers <- c(14, 27, 455)
sum(my_numbers)
```

### What about data that are not numeric?
R can work with categorical data as well as numeric data.  For example, we could create a list of words and store it as a variable if we wanted (feel free to try changing the words if you want):

```{r cat-var, exercise=TRUE}
my_words <- c('RStudio', 'is', 'awesome')
my_words
```

### What if I have a LOT more data to store?
`c` works great for creating small lists of just a few values, but it is not a good way to enter or store large data tables.  In R, these larger datasets are stored as objects called *data.frame*s.  The next sections will get you started using them. 

## How should data tables be organized for statistical analysis?
A comprehensive guide to good practices for formatting data tables is available at <http://kbroman.org/dataorg/>.

A few key points to keep in mind:

- This data table is for the computer to read, not for humans! So eliminate formatting designed to make it "pretty" (colors, shading, fonts...)
- Use short, simple variable names that do not contain any spaces or special characters (like ?, \$, \%, _, etc.)
- Organize the table so there is one column for every variable, and one row for every observation (person/place/thing for which data were collected).
- Use informative variable values rather than arbitrary numeric codes. For example, a variable Color should have values 'red', 'white', and 'blue' rather than 1, 2, and 3.

You will have chances to practice making your own data files and importing them into R outside this tutorial.

## Using built-in datasets in R
  R has a number of built-in datasets that are accessible to you as soon as you start RStudio.
  
  In addition to the datasets that are included with base R, there are add-on *packages* for R that contain additional software tools and sometimes datasets.  
  
  To use datasets contained in a package, you have to load the package by running the command:
  
```{r library-example, eval=FALSE}
library(packagename) 
```

The command `require()` does basically the same thing as `library()`. You can use whichever one you prefer.

```{r require-example, eval=FALSE}
require(packagename)
```

### Example of loading a package
For example, we will practice looking at a dataset from the package `mosaic`. 

Before we can access the data, we have to load the package.  

Edit the code below to load the `mosaic` package and then click "run code" to run it.

**(Nothing obvious will happen...this command just gives R permission to access the package, without asking it to actually DO anything.)**

```{r require-package, echo = FALSE, exercise=TRUE}
library()
```

```{r require-package-hint-1}
library(mosaic)
```

### Viewing a dataset
The mosaic package includes a dataset called `HELPrct`.

If you just enter the dataset name (`HELPrct`) as a command, R will print some of the dataset out to the screen:

```{r}
HELPrct
```

That's not really useful...how can we extract selected, useful information about a dataset?


### Gathering information about a dataset

There are a few functions that make it easier to take a quick look at a dataset:

- `head` prints out the first few rows of the dataset.
- `glimpse` gives an overview of the dataset
- `names` prints out the names of the variables (columns) in the dataset
- `summary` prints out a summary of all the variables (columns) in the dataset
- `nrow` reports the number of rows (observations or cases) in the dataset
- `ncol` reports the number of columns (variables) in the dataset

Try applying each of these functions to the `HELPrct` data and see what the output looks like each time:

```{r check-out-data, exercise=TRUE}
HELPrct
```

```{r check-out-data-hint-1}
head(HELPrct)
```

```{r check-out-data-hint-2}
glimpse(HELPrct)
```

```{r check-out-data-hint-3}
names(HELPrct)
summary(HELPrct)
nrow(HELPrct)
ncol(HELPrct)
```

### Getting more help
You can get help related to R function, and built-in R datasets, using a special function: `?`.  Just type ? followed by the name of the function or dataset you want help on:

```{r get-help, exercise=TRUE}

```

```{r get-help-hint, results='hide'}
?HELPrct
```


## Reading in data from a file
For this class, you will often be asked to analyze data that is stored in files that are available online - usually in csv format. It's simple to read them into R. For example, we can read in the file MI_lead.csv, which is stored at <http://sldr.netlify.com/data/MI_lead.csv>:

```{r}
MI_lead <- read.csv(file='http://sldr.netlify.com/data/MI_lead.csv')
```

### The most common mistakes
The code below contains a couple of the **most common mistakes** students make when they try to read in a data file. See if you can find and correct them.

The code below should read in some baseball statistics from the file <http://sldr.netlify.com/data/baseball.csv>.

```{r baseball-mistakes, exercise = TRUE}
read.csv(http://sldr.netlify.com/data/seball)
```

```{r baseball-mistakes-hint-1}
read.csv(http://sldr.netlify.com/data/baseball)
```

```{r baseball-mistakes-hint-2}
read.csv(http://sldr.netlify.com/data/baseball.csv)
```

```{r baseball-mistakes-hint-3}
read.csv('http://sldr.netlify.com/data/baseball.csv')
```

```{r baseball-mistakes-hint-4}
 <- read.csv('http://sldr.netlify.com/data/baseball.csv')
```

```{r baseball-mistakes-hint-5}
baseball <- read.csv('http://sldr.netlify.com/data/baseball.csv')
```

### What about local files?
The same function, `read.csv()`, can be used to read in a local file.  You just need to change the input to `read.csv()` -- instead of a URL, you provide a path and filename (in quotes).  For example, the input file='http://sldr.netlify.com/data/MI_lead.csv' might become file='C:\\Data\\MI_lead.csv'.

We won't do an example in this tutorial because there isn't a way to work with local files within a tutorial environment, but you can practice it once you are working independently in RStudio.

If you are working on the server rstudio.calvin.edu, you will have to *upload* files to your cloud space on the server before you can read them in (RStudio on the server cannot access files on your computer's hard drive).

### Side note: named input arguments
The input argument we provided to R is the URL (in quotes -- either single or double quotes are fine).  But notice that this time, we gave the input argument a *name*, "file", and specified its value with an equal sign.

This is not *required* - the command works fine without it:

```{r}
MI_lead <- read.csv('http://sldr.netlify.com/data/MI_lead.csv')
```

However, if a function has *more than just one* input argument, it's good to get in the habit of providing names for the inputs.  If you provide names, then the order in which you list the inputs doesn't matter; without names, **the order matters** and you have to use ? to figure out what order R expects!

### Your turn
Use one of the functions you learned earlier to find out how many variables are in the MI_lead dataset, and what their names are. 

```{r MI_lead-info, exercise=TRUE}
```

```{r MI_lead-info-hint, results='hide', echo=FALSE}
names(MI_lead)
```

### Renaming variables in a dataset

This is an advanced topic, so don't worry if it seems complicated; for now, it just nice to realize some of the power R has to clean up and reorganize data.

What if we didn't like the names of the `MI_lead` variables?  For example, a new user of the dataset might not know that that ELL stands for "elevated lead levels" and that ELL2005 gives the *proportion* of tested kids who had elevated lead levels in the year 2005.

If we wanted to use a clearer (though longer) variable name, we might prefer "prop_elevated_lead_2005" instead of "ELL2005" -- more letters to type, but a bit easier to understand for a new user.  How can we tell R we want to rename a variable?

We use the code:

```{r, rename-MI_lead, echo=TRUE}
MI_lead <- MI_lead %>%
  rename(prop_elevated_lead_2005 = ELL2005)

glimpse(MI_lead)
```

The code above uses some tools you've seen, and some more advanced ones you haven't seen yet.  Translated into words, it tells R:

"Make a dataset called `MI_lead` by starting with the dataset `MI_lead`. Send (%>%) the `MI_lead` dataset to the function rename(). What I want to rename is the variable `ELL2005`. Its new name should be `prop_elevated_lead_2005`."

*See...you can already start to make sense of even some pretty complicated (and useful) code.*

### Check out the data
OK, back to business - simple functions and datasets in R.

It's your turn to practice now. Use one of the functions you have learned so far to extract some information about the MI_lead dataset. *Remember, `?` won't work on MI_lead because it's not a built-in R dataset.*

```{r look-at-MI_lead, exercise=TRUE}

```

## Review

What have you learned so far?  More than you think!

### Functions in R
You've learned that R code is made up of functions, which are generally named descriptively according to the job they do.  Functions have one or more input arguments, which is where you provide R with all the data and information it needs to do the job.  The syntax for calling a function uses the template:

  <center>
  <h2><strong><span class="boxed">function</span> ( <span class="boxed">&nbsp;information1&nbsp; , &nbsp;information2&nbsp; , ...)</strong> 
  </h2></center>
  &nbsp;
  
  
### Variables in R
You've practiced creating variables in R using `c`, and saving information (or the results of a computation) using the assignment arrow <-.  

### Datasets in R
You've considered several different ways to get datasets to work with in R: you can use datasets that are built in to R or R packages, or you can use `read.csv` to read in data files stored in .csv format.

## Congratulations!
You just completed your first tutorial on R, and wrote some of your own R code. I *knew* you could do it...

